Optimizing Abstract$^2$ Machines

### 2012-10-14 paper/paper.tex
Setting up definitions.

\begin{figure}
\begin{align*}
\ev{\svar{x},\rho,\kappa,\sigma} &\machstep
\co{\kappa,v,\sigma}
\text{ where }v \in \msto(\menv(x))
\\
\ev{\snum{n},\menv,\mkont,\msto} &\machstep
\co{\mkont,n,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\mkont,\msto} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv}}
\\
%% \ev{\srec\mvarf\mvar\mexp,\menv,\mkont,\msto} &\machstep
%% \co{\mkont,\rlos{\mvarf,\mvar,\mexp,\menv}}
%% \\
\ev{\sapp{\mexpi0}{\mexpi1},\menv,\mkont,\msto} &\machstep
\ev{\mexpi{0},\menv,\kar{\mexpi{1},\menv,a},\msto'}
\text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
\\
\ev{\sif{\mexpi0}{\mexpi1}{\mexpi2},\menv,\mkont,\msto} &\machstep
\ev{\mexpi0,\menv,\kif{\mexpi1,\mexpi2,\menv,\maddr},\msto'}
\text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
\\
\ev{\sopu\mop\mexp,\menv,\mkont,\msto} &\machstep
\ev{\mexp,\menv,\kuop{\mop,\maddr},\msto'}
\text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
%%\\
%% \ev{\sopb\mop{\mexpi0}{\mexpi1},\menv,\mkont,\msto} &\machstep
%% \ev{\mexpi0,\menv,\kbopa{\mexpi1,\menv,\maddr},\msto'}
%% \text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
\\[2mm]
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev{\mexp,\menv,\kfn{\mval,\maddr},\msto}
\\
\co{\kfn{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev{\mexpi0,\menv,\mkont,\msto}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev{\mexpi1,\menv,\mkont,\msto}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kuop{\mop,\maddr},\mval,\msto} & \machstep
\co{\mkont,\mval',\msto}
\text{ where }\mkont\in\msto(\maddr)
\text{ and } \mval'\in\delta(\mop,\mval)
%% \\
%% \co{\kbopa{\mop,\mexp,\menv,\maddr},\mval,\msto} & \machstep
%% \ev{\mexp,\menv,\kbopb{\mop,\mval,\maddr},\msto}
%% \\
%% \co{\kbopb{\mop,\mvalx{u},\maddr},\mval,\msto} & \machstep
%% \co{\mkont,\mval',\msto}
%% \text{ where }\mkont=\msto(\maddr)
%% \text{ and } \mval'=\delta(\mop,\mval,\mvalx{u})
\\[2mm]
\ap{\clos{\mvar,\mexp,\menv},\mval,\mkont} & \machstep
\ev{\mexp,\menv',\mkont,\msto'}
\text{ where }\menv',\msto' = \mathit{bind}(\msto,\mvar,\mval)
\\
%% \ap{\rlos{\mvarf,\mvar,\mexp,\menv},\mval,\mkont} & \machstep
%% \ev{\mexp,\menv',\mkont,\msto'}
%% \text{ where }\menv',\msto' = \mathit{bind}(\msto,\langle\mvar,\mval\rangle,\langle\mvarf,
%% \rlos{\mvarf,\mvar,\mexp,\menv}\rangle)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure}

@@

It is important to recognize from this definition that \emph{any}
allocation strategy is an abstract interpretation.  In particular,
concrete intepretation is a kind of abstract interpretation.  So is an
interpretation that allocates a single cell into which all bindings
and continuations are stored.  On the one hand is an abstract
intepretation that is non-computable and gives only the ground truth
of a programs behavior; on the other is an abstract interpretation
that is easy to compute but gives little information.  Useful program
analyses lay somewhere in between and can be characterized by their
choice of address representation and allocation strategy.
Uniform \(k\)-CFA is one such analysis.

>>>>

ISWIM is a family of programming languages parameterized by a set of
base values and operations.  To make things concrete, we consider a
member of the ISWIM family with integers, booleans, and a few
operations.

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Expressions} & \mathit{e} &=& \svar[^\mlab]\mvar\\
&&|& \slit[^\mlab]\mlit\\
&&|& \slam[^\mlab]\mvar\mexp\\
&&|& \sapp[^\mlab]\mexp\mexp \\
&&|& \sif[^\mlab]\mexp\mexp\mexp \\
\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
\end{array}
\]
\caption{Syntax of ISWIM}
\label{fig:syntax}
\end{figure}

Figure~\ref{fig:syntax} defines the (abstract) syntax of ISWIM.
Figure~\ref{fig:aam} defines the semantics of ISWIM as a machine
model.  Evaluation is defined as the set of states reachable by the
reflexive, transitive closure of the machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply'' form.

Compared with the standard machine semantics, this definition is
different in the following ways:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values,
\item continuations are heap-allocated,
\item there are ``contour values'' (written $\mcntr$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicity parameterized by the functions
  $\mathit{push}$, $\mathit{bind}$, and $\interpdelta$.
\end{itemize}

\begin{figure}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\epsilon}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
\\
\ap{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where }\mkont\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure}

@@

It is important to recognize from this definition that \emph{any}
allocation strategy is an abstract interpretation.  In particular,
concrete intepretation is a kind of abstract interpretation.  So is an
interpretation that allocates a single cell into which all bindings
and continuations are stored.  On the one hand is an abstract
intepretation that is non-computable and gives only the ground truth
of a programs behavior; on the other is an abstract interpretation
that is easy to compute but gives little information.  Useful program
analyses lay somewhere in between and can be characterized by their
choice of address representation and allocation strategy.

%% We now have a framework for describing program analysis for the ISWIM
%% family of languages, whereby approximation of both control and
%% environment structure is regulated by the heap and allocation
%% policies.

Uniform \(k\)-CFA is one such analysis.
